Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-02-17 submission/FILES_TO_ARXIV_16feb20/T1/nlp_models.tex
files I uploaded to arXiv

%KDD% 
\vspace{-1mm}
\section{Comparison of NLP Models}
\label{sxn:nlp}
%\vspace{-1mm}

In this section, we examine empirical quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
%
%In particular, 
Within the past two years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
%
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrices~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger PL exponents $\alpha$.
Based on HT-SR Theory
(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
%
More generally, compared to the CV models of Section~\ref{sxn:cv},
modern NLP models have larger weight matrices and display different spectral properties.
%, both in terms of their architecture (e.g., whether there are convolutions) as well as how they are used (e.g., whether one uses prediction error or perplexity).
Thus, they provide a very different test for our empirical quality metrics.

@@

%KDD% 
\vspace{-1mm}
\paragraph{What do large values of $\alpha$ mean?}

@@

%KDD% 
\vspace{-1mm}
\paragraph{OpenAI GPT Models.}

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
%
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can identify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously,
an example of what we call a series of ``good-better-best'' models.

The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{huggingface}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrices.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
%
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
%
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12$, $24$, $36$, and $48$ layers, respectively, with increasingly larger weight matrices.
%The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
%KDD% \begin{tabular}{|p{0.67in}|c|c|c|c|c|}
\hline
 Series  & \#   & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\hat{\alpha}$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT2-small & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
\hline
GPT2-medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2-large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2-xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for the average Log Norm and Weighted Alpha metrics for pretrained OpenAI GPT and GPT2 models. Column \# refers to number of layers treated.  Note that the averages do not include the first embedding layer(s) because they are not (implicitly) normalized.  }
\label{table:nlp}
\end{table}

%KDD% 
\vspace{-1mm}
\paragraph{Average Quality Metrics for GPT and GPT2.}

@@

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unusually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$.
It also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the ``bad'' GPT versus the ``good'' GPT2-small.
%
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

%KDD% 
\vspace{-1mm}
\paragraph{Scale Collapse in Poorly Trained Models.}

@@

\begin{figure}[h]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-alpha-hist.png}
        \includegraphics[width=5.5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    %KDD% 
    \qquad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-snorm-hist.png}
        \includegraphics[width=5.5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Histogram of PL exponents ($\alpha$) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.}

@@

(Note that Figure~\ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms.
Upon examination, e.g., from Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
%That is, they correspond to the change in normalization of the word embedding layer(s), and thus they can be ignored.
These layers have a different effective normalization, and therefore a different scale.
% \charles{We discuss this further in Appendix~\ref{sxn:appendix}.}
% For example, in GPT, for most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$, but in the first embedding layer, the maximum is of order $N$ (the number of words in the embedding), or $\lambda_{max}\sim\mathcal{O}(10^{5})$.  
% For GPT and GPT2, we treat all layers as-is (although one may want to normalize the first 2 layers by $\mathbf{X}$ by $1/N$ or treat them as an outlier).
We discuss this further in Appendix~\ref{sxn:appendix}.
Here, we do not include them in our computed average metrics in Table~\ref{table:nlp}, and we do not include them in the histogram plot in Figure~\ref{fig:GPT-snorm-hist}.)

%KDD% 
\vspace{-1mm}
\paragraph{Layer Analysis: Correlation Flow and Scale Collapse in GPT and GPT2.}

@@

\begin{figure}[htb]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-alpha-depth.png}
        \includegraphics[width=5.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    %KDD% 
    \quad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT-snorm-depth.png}
        \includegraphics[width=5.5cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{PL exponents ($\alpha$) (in (a)) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) (in (b)) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.  (Note that the quantities being shown on each Y axis are different.)
             In the text, this is interpreted in terms of \emph{Correlation Flow} and \emph{Scale~Collapse}.
            }
    \label{fig:gpt-alpha-layers}
\end{figure}

%KDD% 
\vspace{-1mm}
\paragraph{GPT2: medium, large, xl.}

@@

    \subfigure[PL exponent ($\alpha$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT2*_fnl_alpha_hist.png}
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    %KDD% 
    \quad
    \subfigure[Log Alpha Norm]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{Histogram of PL exponents ($\alpha$)
             %Log Spectral norm ($\log\Vert\mathbf{W}\Vert_{\infty}$), 
              and Log Alpha Norm ($\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$) for weight matrices from models of different sizes in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anomalously large values.)
             }
    \label{fig:gpt2-histograms}
\end{figure}

>>>>

\vspace{-1mm}
\section{Comparison of NLP Models}
\label{sxn:nlp}

In this section, we examine empirical quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
Within the past two years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrices~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger PL exponents $\alpha$.
Based on HT-SR Theory
(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
More generally, compared to the CV models of Section~\ref{sxn:cv},
modern NLP models have larger weight matrices and display different spectral properties.
Thus, they provide a very different test for our empirical quality metrics.

@@

\vspace{-1mm}
\paragraph{What do large values of $\alpha$ mean?}

@@

\vspace{-1mm}
\paragraph{OpenAI GPT Models.}

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can identify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously,
an example of what we call a series of ``good-better-best'' models.

The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{huggingface}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrices.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12$, $24$, $36$, and $48$ layers, respectively, with increasingly larger weight matrices.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
\hline
 Series  & \#   & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\hat{\alpha}$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT2-small & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
\hline
GPT2-medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2-large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2-xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for the average Log Norm and Weighted Alpha metrics for pretrained OpenAI GPT and GPT2 models. Column \# refers to number of layers treated.  Note that the averages do not include the first embedding layer(s) because they are not (implicitly) normalized.  }
\label{table:nlp}
\end{table}

\vspace{-1mm}
\paragraph{Average Quality Metrics for GPT and GPT2.}

@@

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unusually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$.
It also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the ``bad'' GPT versus the ``good'' GPT2-small.
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

\vspace{-1mm}
\paragraph{Scale Collapse in Poorly Trained Models.}

@@

\begin{figure}[h]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    \qquad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        \includegraphics[width=5.5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Histogram of PL exponents ($\alpha$) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.}

@@

(Note that Figure~\ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms.
Upon examination, e.g., from Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
These layers have a different effective normalization, and therefore a different scale.
We discuss this further in Appendix~\ref{sxn:appendix}.
Here, we do not include them in our computed average metrics in Table~\ref{table:nlp}, and we do not include them in the histogram plot in Figure~\ref{fig:GPT-snorm-hist}.)

\vspace{-1mm}
\paragraph{Layer Analysis: Correlation Flow and Scale Collapse in GPT and GPT2.}

@@

\begin{figure}[htb]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \quad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        \includegraphics[width=5.5cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{PL exponents ($\alpha$) (in (a)) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) (in (b)) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.  (Note that the quantities being shown on each Y axis are different.)
             In the text, this is interpreted in terms of \emph{Correlation Flow} and \emph{Scale~Collapse}.
            }
    \label{fig:gpt-alpha-layers}
\end{figure}

\vspace{-1mm}
\paragraph{GPT2: medium, large, xl.}

@@

    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT2xxx_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    \quad
    \subfigure[Log Alpha Norm]{
        \includegraphics[width=5.5cm]{img/GPT2xxx_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{Histogram of PL exponents ($\alpha$)
              and Log Alpha Norm ($\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$) for weight matrices from models of different sizes in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anomalously large values.)
             }
    \label{fig:gpt2-histograms}
\end{figure}

### 2020-02-17 submission/FILES_TO_ARXIV_16feb20/introduction.tex
files I uploaded to arXiv

%KDD% \vspace{-1mm}
\section{Introduction}
\label{sxn:intro}
%\vspace{-1mm}

A common problem in machine learning (ML) 
%and artificial intelligence (AI) 
is to evaluate the quality of a given model.
A popular way to accomplish this
%, in particular in academic environments, 
is to train a model and then evaluate its training/testing error.
There are many problems with this approach.
The training/testing curves give very limited insight into the overall properties of the model; 
they do not take into account the (often large human and CPU/GPU) time for hyperparameter fiddling;
they typically do not correlate with other properties of interest such as robustness or fairness or interpretability; 
and so on.
A less well-known problem, but one that is increasingly important, in particular in industrial-scale artificial intelligence (AI), arises when the model \emph{user} is not the model \emph{developer}.
Here, one may not have access to either the training data or the testing data.
Instead, one may simply be given a model that has already been trained---a \emph{pretrained model}---and need to use it as-is, or to fine-tune and/or compress it and then use it.

@@

In this paper, we present and evaluate quality metrics for pretrained deep neural network (DNN) models, and we do so at scale.
We consider a large suite of hundreds of publicly-available models, mostly from computer vision (CV) and natural language processing (NLP).
%
By now, there are many such state-of-the-art models that are publicly-available, e.g., 
there are now hundreds of pretrained models in CV ($\ge 500$) and NLP ($\approx 100$).%
\footnote{When we began this work in 2018, there were fewer than tens of such models; now in 2020, there are hundreds of such models; and we expect that in a year or two there will be an order of magnitude or more of such models.}
These provide a large corpus of models that by some community standard are state-of-the-art.%
\footnote{Clearly, there is a selection bias or survivorship bias here---people tend not to make publicly-available their poorly-performing models---but these models are things in the world that (like social networks or the internet) can be analyzed for their properties.}
Importantly, all of these models have been trained by someone else and have been viewed to be of sufficient interest/quality to be made publicly-available; and, for all of these models, we have no access to training data or testing data, and we have no knowledge of the training/testing protocols.

@@

%\begin{itemize}[leftmargin=*]
%\item 
%First, 
%Motivated by practical problems, we formulate the question ``Can one predict trends in the quality of state-of-the-art neural networks without access to training or testing data''
%\item
%Second, 
%To answer this question, we analyze hundreds of publicly-available pretrained models, including older and current state-of-the-art models in CV and NLP.
%\end{itemize}

@@

\noindent
We emphasize that our goal is a practical theory to predict trends in the quality of state-of-the-art DNN models, i.e., not to make a statement about every publicly-available model.
We have examined hundreds of models, and we identify general trends, but we also highlight interesting exceptions.
%%Several of the most interesting are described below.

%KDD% \vspace{-1mm}
\paragraph{The WeightWatcher Tool.}

@@

%KDD% \vspace{-1mm}
\paragraph{Organization of this paper.}

>>>>

\section{Introduction}
\label{sxn:intro}

A common problem in machine learning (ML) 
is to evaluate the quality of a given model.
A popular way to accomplish this
is to train a model and then evaluate its training/testing error.
There are many problems with this approach.
The training/testing curves give very limited insight into the overall properties of the model; 
they do not take into account the (often large human and CPU/GPU) time for hyperparameter fiddling;
they typically do not correlate with other properties of interest such as robustness or fairness or interpretability; 
and so on.
A less well-known problem, but one that is increasingly important, in particular in industrial-scale artificial intelligence (AI), arises when the model \emph{user} is not the model \emph{developer}.
Here, one may not have access to either the training data or the testing data.
Instead, one may simply be given a model that has already been trained---a \emph{pretrained model}---and need to use it as-is, or to fine-tune and/or compress it and then use it.

@@

In this paper, we present and evaluate quality metrics for pretrained deep neural network (DNN) models, and we do so at scale.
We consider a large suite of hundreds of publicly-available models, mostly from computer vision (CV) and natural language processing (NLP).
By now, there are many such state-of-the-art models that are publicly-available, e.g., 
there are now hundreds of pretrained models in CV ($\ge 500$) and NLP ($\approx 100$).%
\footnote{When we began this work in 2018, there were fewer than tens of such models; now in 2020, there are hundreds of such models; and we expect that in a year or two there will be an order of magnitude or more of such models.}
These provide a large corpus of models that by some community standard are state-of-the-art.%
\footnote{Clearly, there is a selection bias or survivorship bias here---people tend not to make publicly-available their poorly-performing models---but these models are things in the world that (like social networks or the internet) can be analyzed for their properties.}
Importantly, all of these models have been trained by someone else and have been viewed to be of sufficient interest/quality to be made publicly-available; and, for all of these models, we have no access to training data or testing data, and we have no knowledge of the training/testing protocols.

@@

@@

\noindent
We emphasize that our goal is a practical theory to predict trends in the quality of state-of-the-art DNN models, i.e., not to make a statement about every publicly-available model.
We have examined hundreds of models, and we identify general trends, but we also highlight interesting exceptions.

\paragraph{The WeightWatcher Tool.}

@@

\paragraph{Organization of this paper.}

### 2020-02-17 submission/FILES_TO_ARXIV_16feb20/methods.tex
files I uploaded to arXiv

%KDD% \vspace{-1mm}
\section{Methods}
\label{sxn:methods}
%\vspace{-1mm}

@@

%KDD% \vspace{-1mm}
\paragraph{DNN Empirical Quality Metrics.}

The best performing empirical quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's \emph{Empirical Correlation Matrix}: $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.%

@@

The first two norms are well-known in ML; the last two deserve special mention.
The empirical parameter $\alpha$ is the Power Law (PL) exponent that arises in the recently-developed HT-SR Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
Operationally, $\alpha$ is determined by using the publicly-available \emph{WeightWatcher} tool~\cite{weightwatcher_package} to fit the Empirical Spectral Density (ESD) of $\mathbf{X}$, i.e., a histogram of the eigenvalues, call it $\rho(\lambda)$, to a truncated PL, 
\begin{equation}
\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\le\lambda_{max}.
\end{equation}
%where $\lambda_{max}$ \emph{is} the largest eigenvalue of $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
Each of these quantities is defined for a given layer $\mathbf{W}$ matrix.

@@

The \emph{Weighted Alpha metric} is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
%(and it approximates the average log $\alpha$-Shatten Norm metric),
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total number of layer weight matrices.
The Weighted Alpha metric was introduced previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a limited set of models.

@@

To avoid confusion, let us clarify the relationship between $\alpha$ and $\hat{\alpha}$.  
We fit the ESD of the correlation matrix $\mathbf{X}$ to a truncated PL, parameterized by 2 values: the PL exponent $\alpha$, and the maximum eigenvalue $\lambda_{max}$.  
(Technically, we also need the minimum eigenvalue $\lambda_{min}$, but this detail does not affect our analysis.)
The PL exponent $\alpha$ measures of the amount of correlation in a DNN layer weight matrix $\mathbf{W}$. 
It is valid for $\lambda\le\lambda_{max}$, and it is scale-invariant, i.e., it does not depend on the normalization of $\mathbf{W}$ or $\mathbf{X}$.
The $\lambda_{max}$ is a measure of the size, or scale, of $\mathbf{W}$.
%
Multiplying each $\alpha$ by the corresponding $\log\lambda_{max}$ weighs ``bigger'' layers more, and averaging this product leads to a balanced, Weighted Alpha metric for the entire~DNN.

%KDD% \vspace{-1mm}
\paragraph{Convolutional Layers and Normalization issues.}
There are several technical issues (regarding spectral analysis of convolutional layers and normalization of empirical matrices) that are important for reproducibility of our results.
See Appendix~\ref{sxn:appendix} for a discussion.

>>>>

\section{Methods}
\label{sxn:methods}

@@

\paragraph{DNN Empirical Quality Metrics.}

The best performing empirical quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's \emph{Empirical Correlation Matrix}: $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.

@@

The first two norms are well-known in ML; the last two deserve special mention.
The empirical parameter $\alpha$ is the Power Law (PL) exponent that arises in the recently-developed HT-SR Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
Operationally, $\alpha$ is determined by using the publicly-available \emph{WeightWatcher} tool~\cite{weightwatcher_package} to fit the Empirical Spectral Density (ESD) of $\mathbf{X}$, i.e., a histogram of the eigenvalues, call it $\rho(\lambda)$, to a truncated PL, 
\begin{equation}
\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\le\lambda_{max}.
\end{equation}
Each of these quantities is defined for a given layer $\mathbf{W}$ matrix.

@@

The \emph{Weighted Alpha metric} is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total number of layer weight matrices.
The Weighted Alpha metric was introduced previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a limited set of models.

@@

To avoid confusion, let us clarify the relationship between $\alpha$ and $\hat{\alpha}$.  
We fit the ESD of the correlation matrix $\mathbf{X}$ to a truncated PL, parameterized by 2 values: the PL exponent $\alpha$, and the maximum eigenvalue $\lambda_{max}$.  
(Technically, we also need the minimum eigenvalue $\lambda_{min}$, but this detail does not affect our analysis.)
The PL exponent $\alpha$ measures of the amount of correlation in a DNN layer weight matrix $\mathbf{W}$. 
It is valid for $\lambda\le\lambda_{max}$, and it is scale-invariant, i.e., it does not depend on the normalization of $\mathbf{W}$ or $\mathbf{X}$.
The $\lambda_{max}$ is a measure of the size, or scale, of $\mathbf{W}$.
Multiplying each $\alpha$ by the corresponding $\log\lambda_{max}$ weighs ``bigger'' layers more, and averaging this product leads to a balanced, Weighted Alpha metric for the entire~DNN.

\paragraph{Convolutional Layers and Normalization issues.}
There are several technical issues (regarding spectral analysis of convolutional layers and normalization of empirical matrices) that are important for reproducibility of our results.
See Appendix~\ref{sxn:appendix} for a discussion.
